Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-12 paper.tex
updated with lars version

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light.  After the galactic center, which hosts many
  astrophysical sources, galaxy clusters are the best known target
  class to detect DM annihilation. We predict Virgo and Fornax to be
  the brightest DM sources and find a particularly low CR induced
  background for Fornax. For a given minimum substructure mass as
  given by the DM free-streaming scale, one needs the largest DM halo
  mass to maximize the substructure boost over the smoothly
  distributed annihilation flux: for galaxy clusters as the largest
  collapsed objects, we find a boost factor of more than 1000. Since
  the annihilation flux of substructures is mostly contributed by the
  regions around the virial radius, the resulting surface brightness
  profiles are almost flat. This makes it very challenging to detect
  this flux with imaging air Cherenkov telescopes since their
  sensitivity drops linearly with radius and they typically encounter
  have $5-10$ linear resolution elements across a cluster; the
  gamma-ray satellite Fermi is better suited. Assuming cold dark
  matter with a substructure mass distribution down to an Earth mass
  and using Fermi upper limits, we rule out the leptophilic models in
  their present form in 14 clusters, and limit the boost from
  Sommerfeld enhancement in Fornax to be less than 2 which corresponds
  to a value in the Milky Way of less than 5. Alternatively, if the
  Sommerfeld enhancement factor is realized in Nature, this would
  imply a limiting substructure mass in Fornax of $\mlim >
  10^2\,\msun$ which would be a hint for an unknown form of warm dark
  matter. We find that Fermi three-year data will be able to constrain
  the smallest mass of substructures for supersymmetric benchmark
  models or alternatively constrain their cross sections for a given
  substructure boost. Although the Perseus and Ophiuchus clusters are
  expected to have the largest CR induced flux according to our
  analytic model that is derived from cosmological hydrodynamical
  cluster simulations, their central active galactic nucleus as well
  as the Galactic foreground makes it hard to detect this flux in the
  Fermi band. The next to best targets to detect the CR flux are
  Virgo, Coma, and Norma -- the Fermi upper limits are closing in on
  our predictions and will soon start to constrain the underlying CR
  physics such as shock acceleration efficiencies or CR transport
  properties.
\end{abstract}

@@

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that there are
certain advantages that work in Cavour of this possible target for
gamma-ray detection of DM annihilation.  Galaxy clusters constitute
the most massive objects in our Universe that are forming today. This
causes their DM subhalo mass function to be less affected by tidal
stripping compared to galaxy sized halos that formed long ago.  The
annihilation luminosities of the DM halo component for e.g. the Virgo
cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in Cavour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than galactic haloes \footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}.

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. In Table~\ref{tab:flux_tab} we show
the gamma-ray flux from four promising clusters with a high gamma-ray
yield. We derive the emission from two different DM models and a
hadronic CR model and find that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo \del{\cite{}} as the prime targets for DM observations, Perseus
for CR induced emission and we additionally decide in favor of the
well studied cluster Coma for comparison. While studying these four
clusters in great detail, we complement the analysis by calculating
the expected gamma-ray fluxes of a nearby X-ray flux complete sample
of galaxy clusters from the Rosat all-sky survey (extended HIghest
X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ & $\B_\rmn{sfe} $$^{(1)}$ &  $\B_\rmn{sub} $$^{(2)}$ \\
         & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ 
         & & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Fornax               & $4.8\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $5.6\times10^{-12}$ & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $106$ & $730$  \\
 Coma                 & $1.3\times10^{-8}$  & $8.3\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $44$  & $1310$ \\
 Virgo                & $1.2\times10^{-7}$  & $4.0\times10^{-10}$ & $8.7\times10^{-9}$  
 & \,\,\,\,\,         & $1.3\times10^{-11}$ & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ 
 & $2.1$  & $83$  & $870$  \\
 Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $54$  & $1280$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The boost due to Sommerfeld enhancement.\\
  (2) The boost from substructures have been intergated out to $\rvir$.\\
  (3) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (4) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included.\\
  (5) Gamma-ray flux induced by CR protons.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. Direct detection methods, which are presently
evolving rapidly use the feeble interaction between DM particles of
the galactic halo and nuclei such as Germanium, Sodium, Iodine, Argon
or Xenon to infer the scattering cross section and a rough estimate of
the mass of the DM particle (for the currently most sensitive search,
see \cite{Aprile:2010um}). A characteristic of this method is that it
basically only depends on the local DM density (which is rather well
determined to be around $0.4$ GeV/cm$^3$) and the basic cross section
for DM - nucleus scattering. A disadvantage is that it does benefit
from the two enhancement mechanisms mentioned above, boost from
substructure and/or the Sommerfeld effect. Also, it cannot be excluded
- although it seems at present improbable - that the local halo
structure is such that the solar system happens to be in an underdense
region.

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{Hooper:2010mq}), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

@@

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{Hooper:2008kg,Ahlers:2009ae}). In that case, DM would more
likely be explained by the standard scenario of a WIMP, for example
the thoroughly studied lightest supersymmetric neutralino (for
reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical DM community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/v_{\rm
  cl}$.

@@

\subsubsection{Supersymmetric dark matter}
The most studied models for DM are supersymmetric ones, especially
models where the lightest supersymmetric particle, in most models the
lightest neutralino, is stable. The stability is assured in viable
supersymmetric models due to a discrete symmetry, $R$-parity. This
symmetry is needed from the particle physics point of view to avoid
fast proton decay, and automatically makes the lightest supersymmetric
particle stable, and therefore a good candidate for DM. In addition,
as the coupling to ordinary matter is given by gauge couplings, the
neutralino is automatically a WIMP candidate. Unfortunately, the
breaking of supersymmetry (which has to be there since otherwise,
e.g., there would exist a scalar electron with the same mass as the
electron) means that a large number of essentially free parameters
enter actual calculations.  The neutralino is then a linear
combination of the supersymmetric partners of the U(1) gauge boson,
$\tilde B$, one of the SU(2) gauge bosons, $\tilde W_3$, and the two
Higgs doublets in the Minimal Supersymmetric extension of the Standard
Model (the MSSM), $\tilde H_1^0$ and $\tilde H_2^0$, with mixing
coefficients which depend on the supersymmetric breaking parameters
\cite{1984NuPhB.238..453E}. The parameter space is very large, too
large in general to scan efficiently even with the most powerful
computers. Therefore, often simplifying assumptions are made to limit
the number of free parameters. Even then, a large number of viable
models is found which give a relic density consistent with the WMAP
data. This means that they are indeed very good templates for WIMP DM,
something that explains their popularity besides the subjective
statement that few more plausible models for DM have been put forward
so far.

@@

We use the \ds package \cite{ds} to compute the mass, couplings and
relic density for a given set of parameters.  Actually, we take
advantage of so-called benchmark models, which have been proposed in
different contexts (in particular in \cite{Bringmann:2008kj} giving
predictions for the MAGIC experiment), and which give a good
representation of models with high enough gamma-ray rates to become
the first candidates for DM indirect detection in imaging air
Cherenkov telescopes. We use the following supersymmetric benchmark
models:

@@

The differential energy spectrum for the IB process is more
complicated (see \cite{1989PhLB..225..372B,2008JHEP...01..049B}), but
can be computed using {\sc DarkSUSY}. The differential photon energy
spectra for all our benchmark models are shown in Fig.~\ref{fig:q_DM}.
The peaking at high photon energy and the resulting bump caused by
inner bremsstrahlung is clearly seen for benchmark models $I´$, $J´$,
and $J^*$. Despite the different particle masses and cross sections,
the spectrum of the continuums emission shows a remarkably similar
shape which suggests that a simple rescaling of particle masses
(responsible for the exponential cutoff in the spectrum) and cross
sections (spectral amplitude) should yield a roughly scale-invariant
continuum emission spectrum except for the presence of the final state
emission feature which depends on the details of the specific decay
channels. The right panel of Fig.~\ref{fig:q_DM} shows the cumulative
number of leptons (electrons and positrons) above a given energy for
our different DM models which is of interested for the high-energy IC
emission of those leptons. Only LP models have energetic enough
electrons such that the IC emission is powerful enough to either be
constrained or detected at GeV energies and higher.

>>>>

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light. We predict the Virgo and Fornax clusters to be the
  brightest DM sources and find a particularly low CR induced
  background for Fornax. For a given minimum substructure mass as
  given by the DM free-streaming scale, one needs the largest DM halo
  mass to maximize the substructure boost over the smoothly
  distributed annihilation flux: for galaxy clusters as the largest
  collapsed objects, we find a boost factor of more than 1000. Since
  the annihilation flux of substructures is mostly contributed by the
  regions around the virial radius, the resulting surface brightness
  profiles are almost flat. This makes it very challenging to detect
  this flux with imaging air Cherenkov telescopes since their
  sensitivity drops linearly with radius and they typically have
  $5-10$ linear resolution elements across a cluster; the gamma-ray
  satellite Fermi is better suited. Assuming cold dark matter with a
  substructure mass distribution down to an Earth mass and using Fermi
  upper limits, we rule out the leptophilic models in their present
  form in 14 clusters, and limit the boost from Sommerfeld enhancement
  in Fornax to be less than 2 which corresponds to a value in the
  Milky Way of less than 5. Alternatively, if the Sommerfeld
  enhancement factor is realized in Nature, this would imply a
  limiting substructure mass in Fornax of $\mlim > 10^2\,\msun$ which
  would be a challenge for structure formation. We find that Fermi
  three-year data will be able to constrain the smallest mass of
  substructures for supersymmetric benchmark models or alternatively
  constrain their cross sections for a given substructure boost. The
  Fermi upper limits are closing in on our predictions for the CR flux
  using an analytic model based on cosmological hydrodynamical cluster
  simulations. We will soon start to constrain the underlying CR
  physics such as shock acceleration efficiencies or CR transport
  properties.
\end{abstract}

@@

3. Galaxy clusters. This possibility has been less studied, however we
noted in a previous Letter \cite{2009PhRvL.103r1302P} that there are
certain advantages that work in favour of this possible target for
gamma-ray detection of DM annihilation.  Galaxy clusters constitute
the most massive objects in our Universe that are forming today. This
causes their DM subhalo mass function to be less affected by tidal
stripping compared to galaxy sized halos that formed long ago.  The
annihilation luminosities of the DM halo component for e.g. the Virgo
cluster and the Draco dwarf scales in a way (see
\cite{2009PhRvL.103r1302P}) that the ratio of gamma-ray luminosities
from the smooth components is around 4, in favour of Virgo. In
addition, there may be a further enhancement due to substructure,
which to a large extent should be unaffected by tidal disruption, at
least in the outer regions.  According to a recent estimate
\cite{2011MNRAS.410.2309G}, more massive haloes tend to have a larger
mass fraction in subhalos.  For example, cluster size haloes typically
have 7.5 per cent of the mass within $r_{200}$ in substructures of
fractional mass larger than $10^{-5}$, which is 25 per cent higher
than for galactic haloes.\footnote{All halo masses and length scales in
  this paper are scaled to the currently favored value of Hubble's
  constant, $H_0 = 70\, \rmn{km~s}^{-1}\,\rmn{Mpc}^{-1}$. We define
  the virial mass $\mvir$ and virial radius $\rvir$ as the mass and
  radius of a sphere enclosing a mean density that is 200 times the
  critical density of the Universe $\rho_{\rmn{cr}}$.}

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. For convenience, we summarize already
in Table~\ref{tab:flux_tab} our results for the gamma-ray flux from
four promising clusters with a high gamma-ray yield. We will in the
following derive the emission from two different DM models and a
hadronic CR model to show that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement (SFE)
is small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax and
Virgo as the prime targets for DM observations, Perseus for CR induced
emission and we additionally decide in favor of the well studied
cluster Coma for comparison. While studying these four clusters in
great detail, we complement the analysis by calculating the expected
gamma-ray fluxes of a nearby X-ray flux complete sample of galaxy
clusters from the Rosat all-sky survey (extended HIghest X-ray FLUx
Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & 
$\mvir$ & $\B_\rmn{sfe} $$^{(1)}$ &  $\B_\rmn{sub} $$^{(2)}$ \\
         & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ 
         & & DM-LP $^{(3)}$ & DM-BM-$\Kp$ $^{(4)}$ & CR-$\pi^0$ $^{(5)}$ & $[10^{14}\,\msun]$ &&  \\
 \hline
 Fornax               & $4.8\times10^{-8}$  & $1.3\times10^{-10}$ & $2.2\times10^{-10}$ 
 & \,\,\,\,\,         & $5.6\times10^{-12}$ & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ 
 & $1.0$  & $106$ & $730$  \\
 Coma                 & $1.3\times10^{-8}$  & $8.3\times10^{-11}$ & $4.0\times10^{-9}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ 
 & $13.8$ & $44$  & $1310$ \\
 Virgo                & $1.2\times10^{-7}$  & $4.0\times10^{-10}$ & $8.7\times10^{-9}$  
 & \,\,\,\,\,         & $1.3\times10^{-11}$ & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ 
 & $2.1$  & $83$  & $870$  \\
 Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$  & $7.1\times10^{-11}$ & $1.4\times10^{-8}$  
 & \,\,\,\,\,         & $1.4\times10^{-12}$ & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ 
 & $7.71$ & $54$  & $1280$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The boost due to Sommerfeld enhancement. See Section II A 1.\\
  (2) The boost from substructures have been integrated out to $\rvir$.\\
  (3) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
  (4) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
  model where the boost from substructures is included. See Sections II A 2 and II B 3.\\
  (5) Gamma-ray flux induced by CR protons. See Section V C. 
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

There are three methods for detecting DM candidates which are
presently employed: First, at particle physics accelerators such as
CERN's LHC, one is now entering into a new energy regime which may
allow the production of the heavy, electrically neutral and long-lived
particles which may constitute DM. From these experiments one may get
the first glimpse of the mass scale beyond the standard model where DM
may reside. However, to really show that any of the hypothetical, new
particles created at the LHC is actually the DM, one has to rely on
the two other methods available for the DM search, namely direct and
indirect detection. Direct detection methods, which are presently
evolving rapidly use the feeble interaction between DM particles of
the galactic halo and nuclei such as Germanium, Sodium, Iodine, Argon
or Xenon to infer the scattering cross section and a rough estimate of
the mass of the DM particle (for the currently most sensitive search,
see \cite{Aprile:2010um,Aprile:2011ts}). A characteristic of this method is that it
basically only depends on the local DM density (which is rather well
determined to be around $0.4$ GeV/cm$^3$) and the basic cross section
for DM - nucleus scattering. A disadvantage is that it does not benefit
from the two enhancement mechanisms mentioned above, boost from
substructure and/or the Sommerfeld effect. Also, it cannot be excluded
- although it seems at present improbable - that the local halo
structure is such that the solar system happens to be in an underdense
region.

In indirect detection, on the other hand, in particular in the photon
channel, one searches for products of DM annihilation in the galactic
halo and beyond. Particularly interesting targets are the galactic
center (for a recent possible indication of a signal, see
\cite{2010arXiv1010.2752H}, however, see also \cite{2010arXiv1012.5839B}.), the dwarf galaxies surrounding the MW
\cite{Strigari:2006rd,Essig:2009jx,2010JCAP...01..031S}, galaxy
clusters (the focus of this work)
\cite{Ghigna:1998vn,Lewis:2002mfa,Boyarsky:2006kc,2006A&A...455...21C,2009PhRvL.103r1302P},
and even the cosmological large scale structure
\cite{Bergstrom:2001jj,Ullio:2002pj,Taylor:2002zd,Elsaesser:2004ap,Cuoco:2010jb,Abazajian:2010sq,Abdo:2010dk,Zavala:2011tt}.

@@

A more standard explanation of the PAMELA/Fermi excess would be
pulsars or other supernova remnants (e.g.,
\cite{1995PhRvD..52.3265A,Hooper:2008kg,Ahlers:2009ae}). In that case,
DM would more likely be explained by the standard scenario of a WIMP,
for example the thoroughly studied lightest supersymmetric neutralino
(for reviews, see
\cite{Jungman:1995df,Bergstrom:2000pn,Bertone:2004pz}). We note that
the soon to be launched AMS-02 experiment on the International Space
Station \cite{ams02} may give interesting clues to the origin of the
antimatter excess.

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) caused intense
activity in the theoretical DM community. At first, there was
scepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld enhancement
(SFE). This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the SFE will be smaller than the one expected in a single
galaxy such as the MW, roughly by a factor $v_{\rm MW}/v_{\rm
  cl}$.

@@

\subsubsection{Supersymmetric dark matter}
The most studied models for DM are supersymmetric ones, especially
models where the lightest supersymmetric particle, in most models the
lightest neutralino, is stable. The stability is assured in viable
supersymmetric models due to a discrete symmetry, $R$-parity. This
symmetry is needed from the particle physics point of view to avoid
fast proton decay, and automatically makes the lightest supersymmetric
particle stable, and therefore a good candidate for DM. In addition,
as the coupling to ordinary matter is given by gauge couplings, the
neutralino is automatically a WIMP candidate. Unfortunately, the
breaking of supersymmetry (which has to be there since otherwise,
e.g., there would exist a scalar electron with the same mass as the
electron) means that a large number of essentially free parameters
enter actual calculations.  The neutralino is then a linear
combination of the supersymmetric partners of the U(1) gauge boson,
$\tilde B$, one of the SU(2) gauge bosons, $\tilde W_3$, and the two
Higgs doublets in the Minimal Supersymmetric extension of the Standard
Model (the MSSM), $\tilde H_1^0$ and $\tilde H_2^0$, with mixing
coefficients which depend on the supersymmetric breaking parameters
\cite{1984NuPhB.238..453E}. The parameter space is very large, too
large in general to scan efficiently even with the most powerful
computers. Therefore, often simplifying assumptions are made to limit
the number of free parameters. Even then, a large number of viable
models is found which give a relic density consistent with the WMAP
data. This means that they are indeed very good templates for WIMP DM,
something that explains their popularity besides the subjective
statement that few more well-motivated models for DM have been put forward
so far.

@@

We use the \ds package \cite{ds} to compute the mass, couplings and
relic density for a given set of parameters.  Actually, we take
advantage of so-called benchmark models, which have been proposed in
different contexts (in particular in \cite{Bringmann:2008kj} giving
predictions for imaging air Cherenkov telescopes, like  MAGIC II and CTA), They give a good
representation of models with high enough gamma-ray rates to become
the first candidates for DM indirect detection in imaging air
Cherenkov telescopes. We use the following supersymmetric benchmark
models:

@@

The differential energy spectrum for the IB process is more
complicated (see \cite{1989PhLB..225..372B,2008JHEP...01..049B}), but
can be computed using {\sc DarkSUSY}. The differential photon energy
spectra for all our benchmark models are shown in Fig.~\ref{fig:q_DM}.
The peaking at high photon energy and the resulting bump caused by
inner bremsstrahlung is clearly seen for benchmark models $I´$, $J´$,
and $J^*$. Despite the different particle masses and cross sections,
the spectrum of the continuum emission shows a remarkably similar
shape which suggests that a simple rescaling of particle masses
(responsible for the exponential cutoff in the spectrum) and cross
sections (spectral amplitude) should yield a roughly scale-invariant
continuum emission spectrum except for the presence of the final state
emission feature which depends on the details of the specific decay
channels. The right panel of Fig.~\ref{fig:q_DM} shows the cumulative
number of leptons (electrons and positrons) above a given energy for
our different DM models which is of interested for the high-energy IC
emission of those leptons. Only LP models have energetic enough
electrons such that the IC emission is powerful enough to either be
constrained or detected at GeV energies and higher.

\begin{figure*}
\begin{minipage}{2.0\columnwidth}
 \includegraphics[width=0.49\columnwidth]{figures/fit.ds.flux.eps}
 \includegraphics[width=0.49\columnwidth]{figures/fit.epflux.int.eps}
 \caption{\it Source functions for different dark matter (DM)
   models. Crosses show the simulated data from {\sc DarkSUSY} and the
   solid lines show the fit to the data. LEFT PANEL: normalized
   differential continuum spectra for four different DM benchmark (BM)
   models; $\Ip$ model (red), $\Jp$ (orange), $\Kp$ (green), and $\Js$
   (blue). We use Eq.~(\ref{eq:bm_cont}) to fit the continuum
   spectra. RIGHT PANEL: number of electron and positron per DM
   annihilation above the electron energy $\ee$ for different DM
   models; $\Ip$ BM model (red), $\Jp$ BM model (orange), $\Kp$ BM
   model (green), and $\Js$ BM model (blue), leptophilic (LP) DM
   annihilating into electrons and positrons (purple), and LP DM
   annihilating into muons (black) without considering Sommerfeld
   boosts. We use Eq.~(\ref{eq:bm_elec}) and Eq.~(\ref{eq:lp_elec}) to
   fit the spectra of electrons and positrons from BM and LP models,
   respectively.}
 \label{fig:q_DM}
\end{minipage}
\end{figure*}
